# Barycentric Centroid of Balance


:::{.callout-note}

## Objectives

In this page, we illustrate how to divide the simplex $\mathcal{S}_3$ into three regions ($R_A, R_B, R_C$) so that each region contains a prescribed amount of probability mass ($p_A, p_B, p_C$) from a continuous distribution (here, Dirichlet).

In a nutshell, we use optimal transport theory (particularly Brenier's theory) for transporting mass from a continuous distribution to a discrete one (point masses at the triangle's corners).

The method is performed in two steps:

1. Definition of a transport cost: squared distance between points.
2. Find a map $T$ that moves each point $\boldsymbol{x}$ in the simplex to a vertex $\boldsymbol{u}_i$ (e.g., $(1,0,0)$) while minimizing total transport cost.


The map $T$ is defined via Laguerre cells (power diagrams), which are convex regions where each point is closer (in a weighted sense) to one vertex than to the others.

The obtained regions together form a weighted Voronoi diagram on the simplex. Each region has exactly the desired share of the total probability mass: $\int_{R_i} f(\boldsymbol{x}) dx = p_i$.

:::

```{r package-settings}
#| warning: false
#| message: false
library(tidyverse)
library(transport)
library(MCMCpack)
library(ggtern)

```

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| file: _myfunctions.R
```

$$
\definecolor{wongBlack}{RGB}{0,0,0}
\definecolor{wongGold}{RGB}{230, 159, 0}
\definecolor{wongLightBlue}{RGB}{86, 180, 233}
\definecolor{wongGreen}{RGB}{0, 158, 115}
\definecolor{wongYellow}{RGB}{240, 228, 66}
\definecolor{wongBlue}{RGB}{0, 114, 178}
\definecolor{wongOrange}{RGB}{213, 94, 0}
\definecolor{wongPurple}{RGB}{204, 121, 167}
\definecolor{colGpeZero}{RGB}{127, 23, 14}
\definecolor{colGpeUn}{RGB}{27, 149, 224}
$$

```{r graphs-settings}
#| code-fold: true
#| code-summary: Codes for graphical parameters.
library(extrafont)
loadfonts(device = "pdf", quiet = TRUE)
# col_categ <- c("A" = "#ffdd55", "B" = "#944edf", "C" = "#3fb3b2")
col_categ <- c("#56B4E9", "#D55E00", "#CC79A7")
font_size <- 20
font_family <- "CMU Serif"

path <- "./figs/"
if (!dir.exists(path)) dir.create(path)
```

Let us consider, as in [Chapter -@sec-transp-categ], a categorical variable $x \in \{{\color{colA}A}, {\color{colB}B}, {\color{colC}C}\}$, with group-specific distributions. We draw random samples from a Dirichlet distribution. Let us start with a concentration parameter $\alpha=(2, 5, 3)$. Each sample lies in the 3-dimensional simplex $\mathcal{S}_3$, and the resulting distribution has density $f$ with respect to Lebesgue measure on $\mathcal{S}_3$.


```{r define-samples}
set.seed(123)
n <- 1000
# Concentration parameter:
alpha <-  c(2, 5, 3)
# Draw n samples
samples <- rdirichlet(n, alpha = alpha)
```


We want to find a partition $R_A,R_B,R_C$ of $\mathcal{S}_3$ such that
$$
\int_{R_i}f(\boldsymbol{x})\mathrm{d}\boldsymbol{x}=p_i,~i\in\{A,B,C\}.
$$
We can use (classical) optimal transport theory to find partitions of the simplex that map probability mass from the Dirichlet to the discrete distribution 
$\boldsymbol{p}=(p_A,p_B,p_C)$, in each vertex of the simplex,
$$
\int_{T^{-1}(\boldsymbol{u}_i)}f(\boldsymbol{x})\mathrm{d}\boldsymbol{x}=p_i,~i\in\{A,B,C\},
$$
where $\{\boldsymbol{u}_A,\boldsymbol{u}_B,\boldsymbol{u}_C\}$ are unit vectors, vertices of the $\mathcal{S}_3$ (i.e., $(1,0,0)$, $(0,1,0)$ and $(0,0,1)$).

:::{.callout-note}

Since the initial measures are absolutely continuous with respect to Lebesgue measure, and since both have finite second moment, there exists a unique optimal transport map $T$ (and this map is the gradient of a convex function).

:::

Let us consider for now that $\boldsymbol{p}=(1/2,1/3,1/6))$.

```{r}
# Unit vectors of S_3
vertices <- matrix(c(
  1, 0, 0,  # A
  0, 1, 0,  # B
  0, 0, 1   # C
), byrow = TRUE, ncol = 3)

# source weights
mass_source <- rep(1 / n, n)
# target weights
mass_target <- c(3, 2, 1) / 6

# Cost matrix (squared Euclidean distance)
cost_matrix <- as.matrix(dist(rbind(samples, vertices))^2)
cost_matrix <- cost_matrix[1:n, (n + 1):(n + 3)]

# We assign eah observation to one vertex
# by minimizing the global transport cost, while matching marginals

# Solve the optimal transport plan
ot_plan <- transport::transport(
  a = mass_source, b = mass_target, costm = cost_matrix, 
  method = "shortsimplex"
)

# Assign each sample to a category based on OT plan
assignment <- rep(NA, n)
# mass each source sends to each target
mass_matrix <- matrix(0, nrow = n, ncol = 3)

for (j in 1:nrow(ot_plan)) {
  from <- ot_plan$from[j]
  to <- ot_plan$to[j]
  mass <- ot_plan$mass[j]
  mass_matrix[from, to] <- mass_matrix[from, to] + mass
}

# Assign each source point to the target it contributes the most mass to
assignment <- max.col(mass_matrix, ties.method = "first")

colnames(samples) <- c("A", "B", "C")

samples_dirichlet_p <- 
  as_tibble(samples) |> 
  mutate(category = colnames(samples)[assignment])
samples_dirichlet_p
```


Each region $R_i$ corresponds to the set of points $\boldsymbol{x}\in\mathcal{S}_3$ such that $\boldsymbol{u}_i$ minimizes $\|\boldsymbol{x}-\boldsymbol{u}_i\|^2-\phi_i$ where $\phi_i\in\mathbb{R}$ is a potential offset (weight, determined via dual optimization). This structure defines a power diagram, also known as a Laguerre–Voronoi diagram (or additively weighted Voronoi diagram). These subsets form a weighted Voronoi tessellation (in barycentric space), as shown in @simplex-baryc-centr-bal-example.

```{r}
#| code-fold: true
#| code-summary: Codes to create the Figure.
#| fig-cap: Barycentric centroid of balance, when $\mathcal{D}(\boldsymbol{\alpha})$ with $\boldsymbol{\alpha}=(2, 5, 3)$ and when $\boldsymbol{p}=(3 ,2,1)/6$.
#| label: simplex-baryc-centr-bal-example
ggtern(
  data = samples_dirichlet_p,
  mapping = aes(x = A, y = B, z = C, color = category)
) +
  geom_point(alpha = .8, size = .5) +
  scale_colour_manual(name = "category", values = col_categ) +
  theme_light(base_size = font_size, base_family = font_family) +
  theme(
    strip.background = element_rect(colour = "black", fill = NA),
    strip.text.x = element_text(colour = "black"),
    strip.text.y = element_text(colour = "black"),
    # strip.text = ggtext::element_markdown(),
    text = element_text(family = font_family, size = unit(font_size, "pt")),
    axis.title = element_text(size = rel(.8)),
    tern.axis.arrow.show = TRUE,
    tern.axis.arrow.sep = .16,
    tern.axis.vshift = .09,
    legend.position = "bottom",
    legend.title = element_text(size = .8 * font_size),
    legend.text = element_text(size = .8 * font_size),
    panel.border = element_rect(colour = NA)
  ) +
  theme_hidetitles() +
  guides(colour = guide_legend(override.aes = list(size = 2)))

```

We can, in addition, identify the intersection, i.e., the point where the minimum of the class-wise kernel density estimates is maximized. We do it by considering a grid over which we estimate the density. We create a function, `generate_simplex_grid()`{.R}, to generate a grid on $\mathcal{S}_3$.

```{r define-generate_simplex_grid}
generate_simplex_grid <- function(resolution = 100) {
  grid <- expand_grid(
    A = seq(0, 1, length.out = resolution),
    B = seq(0, 1, length.out = resolution)
  ) |> 
    mutate(C = 1 - A - B) |> 
    filter(C >= 0)
  
  as_tibble(grid)
}
```

We create a function, `get_category_density_2D()`{.R}, to estimate the kernel density estumation for a given category.

```{r define-get_category_density_2D}
get_category_density_2D <- function(samples, 
                                    grid_points,
                                    category_label) {
  
  category_data <- samples |> 
    filter(category == !!category_label) |> 
    dplyr::select("A", "B")
  eval_grid <- grid_points[, c("A", "B")]
  
  if (nrow(category_data) < 10) {
    H <- diag(0.01, 2)
  } else {
    H <- tryCatch(Hpi(category_data), error = function(e) diag(0.01, 2))
  }
  
  kde_result <- ks::kde(x = category_data, H = H, eval.points = eval_grid)
  
  kde_result$estimate
}
```

Let us create a triangular grid over (A,B,C) constrained to S_3:
```{r define-grid_points}
grid_points <- generate_simplex_grid(resolution = 100)
```

We evaluate the KDE for data from the source distribution:
```{r define-dens_A}
dens_A <- get_category_density_2D(
  samples = samples_dirichlet_p, grid_points = grid_points, category_label = "A"
)
dens_B <- get_category_density_2D(
  samples = samples_dirichlet_p, grid_points = grid_points, category_label = "B"
)
dens_C <- get_category_density_2D(
  samples = samples_dirichlet_p, grid_points = grid_points, category_label = "C"
)
```

Then, we can find the point where `min(densities)`{.R} is maximal.

```{r define-max_idx}
min_dens <- pmin(dens_A, dens_B, dens_C)
max_idx <- which.max(min_dens)
intersection_point <- grid_points[max_idx, ]
tb_intersection <- as_tibble(intersection_point)
```


```{r}
#| code-fold: true
#| code-summary: Codes to create the Figure.
#| fig-cap: "Barycentric centroid of balance, when $\\mathcal{D}(\\boldsymbol{\\alpha})$ with $\\boldsymbol{\\alpha}=(2, 5, 3)$ and when $\\boldsymbol{p}=(3,2,1)/6$. Black dot: intersection (point where the minimum of the class-wise kernel density estimates is maximized)."
#| label: fig-simplex-baryc-centr-bal-example

p <- ggtern(
    data = samples_dirichlet_p,
  mapping = aes(x = A, y = B, z = C)
) +
  geom_point(alpha = .8, size = .5, mapping = aes(color = category)) +
  geom_point(data = tb_intersection) +
  scale_colour_manual(values = col_categ) +
  theme_light(base_size = font_size, base_family = font_family) +
  theme(
    strip.background = element_rect(colour = "black", fill = NA),
    strip.text.x = element_text(colour = "black"),
    strip.text.y = element_text(colour = "black"),
    text = element_text(family = font_family, size = unit(font_size, "pt")),
    axis.title = element_text(size = rel(.8)),
    tern.axis.arrow.show = TRUE,
    tern.axis.arrow.sep = .16,
    tern.axis.vshift = .09,
    legend.position = "bottom",
    legend.title = element_text(size = .8 * font_size),
    legend.text = element_text(size = .8 * font_size),
    panel.border = element_rect(colour = NA)
  ) +
  theme_hidetitles() +
  guides(colour = guide_legend(override.aes = list(size = 2)))

p
```

## Varying $\alpha$ and $\boldsymbol{p}$.


Let us now make $\alpha$ and $\boldsymbol{p}$ vary. We will consider $\boldsymbol{\alpha} = (1,1,1)$ (uniform distribution), and $\boldsymbol{\alpha} = (2, 5, 3)$;  and $\boldsymbol{p} = (1/3,1/3,1/3)$, and $\boldsymbol{p} = (1/2,1/3,1/6)$.

For convenience, let us wrap the previous code in a function, `get_data_assignment()`{.R}.

```{r define-get_data_assignment}
#| code-fold: true
#| code-summary: Code for the `get_data_assignment`{.R} function
#' @param n Number of observations to sample from the Dirichlet Distribution.
#' @param Vector of shape parameters, or matrix of shape parameters 
#'  corresponding to the number of draw. Default to \eqn{(1,1,1)}.
#' @param p Vector of target probabilities. Default to \eqn{(1/3, 1/3, 1/3)}.
#' 
get_data_assignment <- function(n,
                                alpha = c(1, 1, 1),
                                p = c(1, 1, 1) / 3,
                                intersection_point = TRUE) {
  
  # Draw n samples
  samples <- rdirichlet(n, alpha = alpha)
  
  # Unit vectors of S_3
  vertices <- matrix(c(
    1, 0, 0,  # A
    0, 1, 0,  # B
    0, 0, 1   # C
  ), byrow = TRUE, ncol = 3)
  
  # source weights
  mass_source <- rep(1 / n, n)
  # target weights
  mass_target <- p
  
  # Cost matrix (squared Euclidean distance)
  cost_matrix <- as.matrix(dist(rbind(samples, vertices))^2)
  cost_matrix <- cost_matrix[1:n, (n + 1):(n + 3)]
  
  # We assign eah observation to one vertex
  # by minimizing the global transport cost, while matching marginals
  
  # Solve the optimal transport plan
  ot_plan <- transport::transport(
    a = mass_source, b = mass_target, costm = cost_matrix, 
    method = "shortsimplex"
  )
  
  # Assign each sample to a category based on OT plan
  assignment <- rep(NA, n)
  # mass each source sends to each target
  mass_matrix <- matrix(0, nrow = n, ncol = 3)
  
  for (j in 1:nrow(ot_plan)) {
    from <- ot_plan$from[j]
    to <- ot_plan$to[j]
    mass <- ot_plan$mass[j]
    mass_matrix[from, to] <- mass_matrix[from, to] + mass
  }
  
  # Assign each source point to the target it contributes the most mass to
  assignment <- max.col(mass_matrix, ties.method = "first")
  
  colnames(samples) <- c("A", "B", "C")
  samples <- 
    as_tibble(samples) |> 
    mutate(category = colnames(samples)[assignment])
  
  #
  # Intersection point
  #
  if (intersection_point == TRUE) {
    # Create a triangular grid over (A,B,C) constrained to S_3
    grid_points <- generate_simplex_grid(resolution = 100)
    
    # Evaluation of KDE for data from the source distribution
    dens_A <- get_category_density_2D(
      samples = samples, grid_points = grid_points, category_label = "A"
    )
    dens_B <- get_category_density_2D(
      samples = samples, grid_points = grid_points, category_label = "B"
    )
    dens_C <- get_category_density_2D(
      samples = samples, grid_points = grid_points, category_label = "C"
    )
    # Find point where min(densities) is maximal
    min_dens <- pmin(dens_A, dens_B, dens_C)
    max_idx <- which.max(min_dens)
    
    intersection_point <- grid_points[max_idx, ]
    
    tb_intersection <- as_tibble(intersection_point)
  } else {
    tb_intersection <- NULL
  }
  
  list(
    samples = samples,
    tb_intersection = tb_intersection
  )
}
```

Using `get_data_assignment()`{.R}, we draw samples according to $\boldsymbol{\alpha}$ and then we find partitions of the simplex that map probability mass from the Dirichlet to the discrete distribution $\boldsymbol{p}$, in each vertex of the simplex.


```{r define-samples_unif_unif}
samples_unif_unif <- get_data_assignment(
  n = n, 
  alpha = c(1, 1, 1), 
  p = c(1, 1, 1) / 3
)

samples_unif_p <- get_data_assignment(
  n = n, 
  alpha = c(1, 1, 1), 
  p = c(3, 2, 1) / 6
)

samples_dirichlet_unif <- get_data_assignment(
  n = n, 
  alpha = c(2, 5, 3), 
  p = c(1, 1, 1) / 3
)

samples_dirichlet_p <- get_data_assignment(
  n = n, 
  alpha = c(2, 5, 3), 
  p = c(3, 2, 1) / 6
)
```

The results can be visualized in simplex-baryc-centr-bal-example-full, when $\mathcal{D}(\boldsymbol{\alpha})$ with $\boldsymbol{\alpha}=(1, 1, 1)$ (left) and $\boldsymbol{\alpha}=(2, 5, 3)$ (right), when $\boldsymbol{p}=(1,1,1)/3$ (top) and $\boldsymbol{p}=(3,2,1)/6$ (bottom).


```{r}
#| code-fold: true
#| code-summary: Codes to create the Figure.
#| fig-cap: "Barycentric centroid of balance, when $\\mathcal{D}(\\boldsymbol{\\alpha})$ with $\\boldsymbol{\\alpha}=(1, 1, 1)$ (left) and $\\boldsymbol{\\alpha}=(2, 5, 3)$ (right), when $\\boldsymbol{p}=(1,1,1)/3$ (top) and $\\boldsymbol{p}=(3,2,1)/6$ (bottom). Black dot: intersection (point where the minimum of the class-wise kernel density estimates is maximized)."
#| label: fig-simplex-baryc-centr-bal-example-full
#| fig-width: 6
#| fig-height: 6
p <- ggtern(
  data = samples_unif_unif$samples |> 
    mutate(distrib = "Uniform", p = "(1,1,1)/3") |> 
    bind_rows(
      samples_unif_p$samples |> 
        mutate(distrib = "Uniform", p = "(3,2,1)/6")
    ) |> 
    bind_rows(
      samples_dirichlet_unif$samples |> 
        mutate(distrib = "2_5_3", p = "(1,1,1)/3")
    ) |> 
    bind_rows(
      samples_dirichlet_p$samples |> 
        mutate(distrib = "2_5_3", p = "(3,2,1)/6")
    ) |>
    mutate(
      distrib = factor(
        distrib,
        levels = c("Uniform", "2_5_3"),
        labels = c(
          "Uniform" = parse(text = latex2exp::TeX("D(1,1,1)")),
          "2_5_3" = parse(text = latex2exp::TeX("$D(2,5,3)$"))
        )
      ),
      p = factor(
        p,
        levels = c("(1,1,1)/3", "(3,2,1)/6"),
        labels = c(
          "(1,1,1)/3" = parse(text = latex2exp::TeX("$p=(1,1,1)/3$")),
          "(3,2,1)/6" = parse(text = latex2exp::TeX("$p=(3,2,1)/6$"))
        )
      )
    ),
  mapping = aes(x = A, y = B, z = C)
) +
  geom_point(alpha = .8, size = .5, mapping = aes(color = category)) +
  # geom_point(
  #   data = samples_unif_unif$tb_intersection |> 
  #     mutate(distrib = "Uniform", p = "(1,1,1)/3") |> 
  #     bind_rows(
  #       samples_unif_p$tb_intersection |> 
  #         mutate(distrib = "Uniform", p = "(3,2,1)/6")
  #     ) |> 
  #     bind_rows(
  #       samples_dirichlet_unif$tb_intersection |> 
  #         mutate(distrib = "2_5_3", p = "(1,1,1)/3")
  #     ) |> 
  #     bind_rows(
  #       samples_dirichlet_p$tb_intersection |> 
  #         mutate(distrib = "2_5_3", p = "(3,2,1)/6")
  #     ) |>
  #     mutate(
  #       distrib = factor(
  #         distrib,
  #         levels = c("Uniform", "2_5_3"),
  #         labels = c(
  #           "Uniform" = parse(text = latex2exp::TeX("D(1,1,1)")),
  #           "2_5_3" = parse(text = latex2exp::TeX("$D(2,5,3)$"))
  #         )
  #       ),
  #       p = factor(
  #         p,
  #         levels = c("(1,1,1)/3", "(3,2,1)/6"),
  #         labels = c(
  #           "(1,1,1)/3" = parse(text = latex2exp::TeX("$p=(1,1,1)/3$")),
  #           "(3,2,1)/6" = parse(text = latex2exp::TeX("$p=(3,2,1)/6$"))
  #         )
  #       )
  #     )
  # ) +
  scale_colour_manual(values = col_categ) +
  facet_grid(p ~ distrib, labeller = label_parsed, switch = "y") +
  theme_light(base_size = font_size, base_family = font_family) +
  theme(
    strip.background = element_rect(colour = "black", fill = NA),
    strip.text.x = element_text(colour = "black"),
    strip.text.y = element_text(colour = "black"),
    text = element_text(family = font_family, size = unit(font_size, "pt")),
    axis.title = element_text(size = rel(.8)),
    tern.axis.arrow.show = TRUE,
    tern.axis.arrow.sep = .16,
    tern.axis.vshift = .09,
    legend.position = "bottom",
    legend.title = element_text(size = .8 * font_size),
    legend.text = element_text(size = .8 * font_size),
    panel.border = element_rect(colour = NA)
  ) +
  theme_hidetitles() +
  guides(colour = guide_legend(override.aes = list(size = 2)))

p
```


```{r, eval=FALSE}
#| code-fold: true
#| code-summary: Codes to export the figure in PDF.
filename <- "baryc-centr-bal"
ggsave(
  p, file = str_c(path, filename, ".pdf"),
  height = 3.3*1.75, width = 3.25*1.75,
  family = font_family,
  device = cairo_pdf
)
# Crop PDF
system(paste0("pdfcrop ", path, filename, ".pdf ", path, filename, ".pdf"))
```

