# Random Matching {#sec-transp-categ-matching}


:::{.callout-note}

## Objectives

We show how random matching between individuals from two groups can be made to assign a counterfactual value for a categorical variable.

This method is not that used in our methodology, but it (hopefully) helps understanding how our approach differs.

:::

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| file: _myfunctions.R
```


```{r}
#| warning: false
#| message: false
#| code-fold: true
#| code-summary: Codes for graphical parameters
# col_categ <- c("#ffdd55","#944edf","#3fb3b2")
col_group <- c("#00A08A","#F2AD00", "#1b95e0")
col_categ <- c("#56B4E9", "#D55E00", "#CC79A7")
colA <- col_categ[1] ; colB <- col_categ[2] ; colC <- col_categ[3]
colGpe1 <- col_group[2]
colGpe0 <- col_group[1]
font_size <- 20
font_family <- "CMU Serif"

path <- "./figs/"

theme_ggtern_paper <- function(...) {
  font_family <- "CMU Serif"
  font_size <- 10
  theme(
    strip.background = element_rect(colour = "black", fill = NA),
    strip.text.x = element_text(colour = "black"),
    strip.text = ggtext::element_markdown(),
    text = element_text(family = font_family, size = unit(font_size, "pt")),
    axis.title = element_text(size = rel(1)),
    tern.axis.arrow.show = TRUE,
    tern.axis.arrow.sep = .13,
    tern.axis.vshift = .05,
    panel.border = element_rect(colour = NA)
  )
}

theme_ggtern <- function(...) {
  font_family <- "CMU Serif"
  font_size <- 20
  theme(
    strip.background = element_rect(colour = "black", fill = NA),
    strip.text.x = element_text(colour = "black"),
    strip.text = ggtext::element_markdown(),
    text = element_text(family = font_family, size = unit(font_size, "pt")),
    axis.title = element_text(size = rel(1)),
    tern.axis.arrow.show = TRUE,
    tern.axis.arrow.sep = .13,
    tern.axis.vshift = .05,
    panel.border = element_rect(colour = NA)
  )
}

theme_ggtern_minimal <- function(...) {
  font_family <- "CMU Serif"
  font_size <- 10
  theme(
    strip.background = element_rect(colour = "black", fill = NA),
    # strip.text.x = element_text(colour = "black"),
    # strip.text = ggtext::element_markdown(),
    text = element_text(family = font_family, size = unit(font_size, "pt")),
    axis.title = element_text(size = rel(1)),
    tern.axis.arrow.show = FALSE,
    tern.axis.arrow.sep = .13,
    tern.axis.vshift = .05,
    panel.border = element_rect(colour = NA),
    tern.axis.text.T = element_blank(),
    tern.axis.text.L = element_blank(),
    tern.axis.text.R = element_blank(),
    tern.axis.vshift = 0.2
  )
}


#' Theme for ggplot2
#'
#' @param ... Arguments passed to the theme function.
#' @export
#' @importFrom ggplot2 element_rect element_text element_blank element_line unit
#'   rel theme
#'
theme_paper <- function (...) {
  theme(
    text = element_text(family = font_family),
    plot.background = element_rect(fill = "transparent", color = NA),
    panel.background = element_rect(fill = "transparent", color = NA),
    # panel.border = element_rect(fill = NA, colour = "black", linewidth = 1),
    panel.border = element_blank(),
    # axis.line = element_line(color = "black"),
    axis.line <- element_blank(),
    axis.text = element_text(color = "black"),
    legend.text = element_text(size = rel(1)),
    legend.title = element_text(size = rel(1)),
    legend.background = element_rect(fill = "transparent", color = NULL),
    # legend.position = "bottom",
    # legend.direction = "horizontal",
    # legend.box = "vertical",
    legend.key = element_blank(),
    panel.spacing = unit(1, "lines"),
    panel.grid.major = element_line(colour = "grey90"),
    panel.grid.minor = element_blank(),
    plot.title = element_text(hjust = 0, size = rel(1), face = "bold"),
    plot.title.position = "plot",
    plot.margin = unit(c(1, 1, 1, 1), "lines"),
    strip.background = element_rect(fill = NA, colour = NA),
    strip.text = element_text(size = rel(1))
  )
}
```


$$
\definecolor{wongBlack}{RGB}{0,0,0}
\definecolor{wongGold}{RGB}{230, 159, 0}
\definecolor{wongLightBlue}{RGB}{86, 180, 233}
\definecolor{wongGreen}{RGB}{0, 158, 115}
\definecolor{wongYellow}{RGB}{240, 228, 66}
\definecolor{wongBlue}{RGB}{0, 114, 178}
\definecolor{wongOrange}{RGB}{213, 94, 0}
\definecolor{wongPurple}{RGB}{204, 121, 167}
%\definecolor{colA}{RGB}{255, 221, 85}
%\definecolor{colB}{RGB}{148, 78, 223}
%\definecolor{colC}{RGB}{63, 179, 178}
\definecolor{colA}{RGB}{0, 114, 178}
\definecolor{colB}{RGB}{213, 94, 0}
\definecolor{colC}{RGB}{204, 121, 167}
\definecolor{colGpeZero}{RGB}{0,160,138}
\definecolor{colGpeUn}{RGB}{242, 173, 0}
$$

```{r}
#| echo: true
#| warning: false
#| message: false
library(dplyr)
library(tidyr)
library(stringr)
library(ggtern)
library(compositions)
library(ggalluvial)
```

Recall from [Chapter -@sec-transp-categ-objectives] that $a\in\{0,1\}$ denotes a binary treatment. Let $x\in\mathcal{X}$ denote a mediator variable. The objective here is to obtain counterfactuals $(a=1,\boldsymbol{x}(1))$ for `r colorize("untreated units", colGpe0)` $(t=0,\boldsymbol{x})$, when the variable $x$ is **categorical**.


As in [Chapter -@sec-transp-categ-objectives], consider the case where $x\in\{A,B,C\}$ with the following group-wise proportions: $\boldsymbol{p}_0 = (0.1, 0.5, 0.4)$ (`r colorize("untreated units", colGpe0)`) and $\boldsymbol{p}_1 = (0.5, 0.3, 0.2)$ (`r colorize("treated units", colGpe1)`). The proportion in `r colorize("group 0", colGpe0)` are shown in @fig-transp-categ-steps-prop0.


## Preliminary Example: 1-to-1 Random Matching

Let us generate a dummy data set with 100 individuals in both `r colorize("Group 0", colGpe0)` and `r colorize("Group 1", colGpe1)`.
```{r}
set.seed(1234)
n <- 100
n0 <- n1 <- n
p0 <- c(0.1, 0.5, 0.4)
p1 <- c(0.5, 0.3, 0.2)
# Sample category
x0 <- sample(c(rep("A", p0[1] * n0), rep("B", p0[2] * n0), rep("C", p0[3] * n0)), replace = FALSE)
x1 <- sample(c(rep("A", p1[1] * n1), rep("B", p1[2] * n1), rep("C", p1[3] * n1)), replace = FALSE)
cat_levels <- c("A", "B", "C")
```


A way to obtain the counterfactual for the categorical variable is to implement a **1-to-1 matching procedure**. 

Each category is assigned an arbitrary numeric value (e.g., $A = 1$, $B = 2$, $C = 3$), allowing us to define a cost matrix based on the absolute difference between encoded categories. That is, the cost of matching an individual from `r colorize("Group 0", colGpe1)` with category $x_{j0}$ to an individual from `r colorize("Group 1", colGpe0)` with category $x_{i1}$ is given by $C_{ij} = |x_{i1} - x_{j0}|$.

A linear sum assignment problem can then be used to find the matching that minimizes the total cost across pairs.

The matched category from `r colorize("Group 1", colGpe1)` is interpreted as the counterfactual category for the initial `r colorize("Group 0", colGpe0)` individual.


We can compute the distance between the observations from `r colorize("Group 0", colGpe0)` to `r colorize("Group 1", colGpe1)`, by setting numeric values to each category: A=1, B=2, C=3:

```{r compute-cost_matrix}
x0_index <- match(x0, cat_levels)
x1_index <- match(x1, cat_levels)
cost_matrix <- outer(x0_index, x1_index, function(i, j) abs(i - j))
```

The linear sum assignment problem is tackled with `solve_LSAP()`{.R} from {clue}.

```{r define-assignment}
library(clue)
assignment <- solve_LSAP(cost_matrix)
```

The mapping can be stored in a tibble.
```{r define-tb_coupling}
tb_coupling <- tibble(
  x0 = x0,
  x1 = x1[assignment]
) |> 
  mutate(
    cost = abs(match(x0, cat_levels) - match(x1, cat_levels))
  )
```

We compute the number of observations from `r colorize("Group 0", colGpe0)` matched with observations from `r colorize("Group 1", colGpe1)` per category.
```{r}
tb_coupling |> 
  group_by(x0, x1) |> 
  count() |> 
  group_by(x0) |> 
  mutate(prop_0 = 100 * n / sum(n)) |> 
  group_by(x1) |> 
  mutate(prop_1 = 100 * n / sum(n))
```

Then, we can visualize the results, using an alluvial plot (@fig-categ-alluv). All individuals in `r colorize("Group 0", colGpe0)` with label `r colorize("A", colA)` are matched directly to individuals in `r colorize("Group 1", colGpe1)` with the same label. The remaining individuals in `r colorize("Group 1", colGpe1)` labeled `r colorize("A", colA)` are then matched to `r colorize("Group 0", colGpe0)` individuals who originally had label `r colorize("B", colB)` or `r colorize("C", colC)`. This matching is performed probabilistically: among `r colorize("Group 0", colGpe0)` individuals with label `r colorize("B", colB)`, 60% retain their original label, and 40% are reclassified as `r colorize("A", colA)`; among those with label `r colorize("C", colC)`, 50% remain `r colorize("C", colC)`, and 50% are reclassified as `r colorize("A", colA)`.

```{r}
#| label: fig-categ-alluv
#| fig-cap: Matching individuals given a categorical variable.
#| code-fold: true
#| code-summary: Codes to create the Figure.
#| warning: false
flow <- data.frame(
  depart = rep(LETTERS[1:3], 3),
  category = rep(LETTERS[1:3], each = 3),
  freq = as.vector(table(tb_coupling$x0, tb_coupling$x1))
)
p_1 <- ggplot(
  data = flow,
  mapping = aes(axis1 = depart, axis2 = category, y = freq)
) +
  geom_alluvium(aes(fill = category)) +
  geom_stratum() +
  scale_fill_manual(values = col_categ) +
  geom_text(
    stat = "stratum",
    mapping = aes(label = after_stat(stratum)),
    family = font_family
  ) +
  # scale_x_discrete(
  #   limits = c("Group 1", "Group 0"),
  #   expand = c(0.15, 0.05)
  # ) +
  scale_x_discrete(
    limits = c("Group 0", "Group 1"),
    labels = c(
      "Group 0" = str_c("<span style='color:", col_group[1], ";'>Group 0</span>"),
      "Group 1" = str_c("<span style='color:", col_group[2], ";'>Group 1</span>")
    ),
    expand = c(0.15, 0.05)
  ) +
  ylab("Proportions") + 
  scale_y_continuous(transform = ) +
  # theme_minimal(base_size = font_size, base_family = font_family) +
  theme_paper() +
  theme(
    axis.text.x = ggtext::element_markdown()
  )
p_1
```


This can also be visualized on a ternary plot (@fig-categ-ternary).

```{r}
#| fig-cap: Matching individuals given a categorical variable, on a Ternary plot.
#| label: fig-categ-ternary
#| code-fold: true
#| code-summary: Codes to create the Figure.
#| warning: false
# Create interpolated values using McCann (1997) displacement
f_line_simplex <- function(x, 
                           y, 
                           lgt = 601) {
  
  zx <- as.numeric(clr(x))[1:2]
  zy <- as.numeric(clr(y))[1:2]
  t <- seq(0, 1, length = lgt)
  
  tx <- cbind(
    (1 - t) * zx[1] + t * zy[1], 
    (1 - t) * zx[2] + t * zy[2]
  )
  tx <- cbind(tx, -(tx[, 1] + tx[, 2]))
  df <- as.data.frame(matrix(as.numeric(clrInv(tx)), lgt, 3))
  names(df) <- c("A","B","C")
  
  df
}

# dummy dataset to create an empty ternary plot
SB <- tibble(
  A = c(0.2, 0.3, 0.5, 0.6),
  B = c(0.3, 0.4, 0.2, 0.1),
  C = 1 - c(0.2, 0.3, 0.5, 0.6) - c(0.3, 0.4, 0.2, 0.1),
  group = c("0", "0", "1", "1")
)

p_2 <- ggtern(data = SB, aes(x = A, y = B, z = C)) +
  scale_colour_manual(name = "group", values = col_group) +
  guides(
    colour = guide_legend(
      override.aes = list(
        linetype = "solid",
        shape = NA,
        size = 1.5,
        alpha = 1
      )
    )
  ) +
  theme_light(base_size = font_size, base_family = font_family) +
  theme_ggtern_paper() +
  theme(
    legend.title = element_text(size = font_size),
    legend.text = element_text(size = font_size)
    # tern.axis.hshift = .10
  ) +
  theme_latex(TRUE) +
  theme_hidetitles()


p_2 <- p_2 + 
  geom_text(mapping = aes(x = 0.9, y = 0.06, z = 0.08), label = p1[1], color = col_group[2], family = font_family, size = font_size-3, size.unit = "pt") +
  geom_text(mapping = aes(x = 0.09, y = 0.9, z = 0.09), label = p1[2], color = col_group[2], family = font_family, size = font_size-3, size.unit = "pt") +
  geom_text(mapping = aes(x = 0.08, y = 0.06, z = 0.9), label = p1[3], color = col_group[2], family = font_family, size = font_size-3, size.unit = "pt") + 
  geom_text(mapping = aes(x = 0.3, y = 0.1, z = 0.11), label = p0[1], color = col_group[1], family = font_family, size = font_size-3, size.unit = "pt") +
  geom_text(mapping = aes(x = 0.15, y = 0.65, z = 0.25), label = p0[2], color = col_group[1], family = font_family, size = font_size-3, size.unit = "pt") +
  geom_text(mapping = aes(x = 0.1, y = 0.2, z = 0.8), label = p0[3], color = col_group[1], family = font_family, size = font_size-3, size.unit = "pt") 


Li1 <- f_line_simplex(x = c(.75, .125, .125), y = c(.125, .125, .75), lgt = 2)
Li2 <- f_line_simplex(x = c(.75, .125, .125), y = c(.125, .75, .125), lgt = 2)
p_2 <- p_2 + 
  geom_line(
    data = Li2, aes(x = A, y = B, z = C), 
    color = col_group[1], linwidth = .6,
    arrow = arrow(length=unit(0.20,"cm"))
  ) + 
  geom_line(
    data = Li1, aes(x = A, y = B, z = C), 
    color = col_group[1], linwidth = .6,
    arrow = arrow(length=unit(0.20,"cm"))
  ) 
p_2
```


```{r}
#| eval: false
#| code-fold: true
#| code-summary: Codes to export the figures in PDF.

p_matching_indiv <- cowplot::plot_grid(
  ggplotGrob(
    p_1 +
      # Remove top/bottom margin
      theme(
        plot.margin = ggplot2::margin(t = 0, r = 0, b = 0, l = 0)
      )
  ),
  # table_grob,
  ggplotGrob(
    p_2 +
      # Remove top/bottom margin
      theme(
        plot.background = element_rect(fill = "transparent", color = NA),
        plot.margin = ggplot2::margin(t = 0, r = 0, b = 0, l = 0)
      )
  ),
  rel_widths = c(1.4,1),
  ncol = 2
)

p_matching_indiv

filename <- "ternary-categ-matching-indiv"
ggsave(
  p_matching_indiv, file = str_c(path, filename, ".pdf"),
  height = 2*1.75, width = 3.75*1.75,
  family = font_family,
  device = cairo_pdf
)
# Crop PDF
system(paste0("pdfcrop ", path, filename, ".pdf ", path, filename, ".pdf"))
```


Rather than randomly matching individuals, we can use optimal transport to do the assignment. In [Chapter -@sec-transp-categ-discrete], we begin by arbitrarily assigning numerical values to each category and computing pairwise distances based on these values. In contrast, [Chapter -@sec-transp-categ-transport] relies on an which consists in representing the categorical variable in the probability simplex, using probability vectors that reflect the likelihood of each category.